An Event-based Architecture for Multi-population Optimization Algorithms

### 2020-01-30 draft.tex
Explaining away results

We present the results of the experiments in this section. The effect of a
different number of populations on execution time is reported in  Section
\ref{sec:exp1results}.
% The results of the performance comparison between homogeneous and
% heterogeneous search strategies are presented in Section \ref{sec:exp2results}.

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{sp_size_worker}
    \caption{Time required, in seconds, to complete an instance of the function, with different number of workers and populations sizes }
    \label{fig:spworker}
\end{figure}

% \subsection{Experiment 1}
% \label{sec:exp1results}

We summarize the results of the experiment in Figure \ref{fig:spworker}.  The
experiment consisted of executing the same benchmark on EvoSwarm twice but
changing the number of populations used in each run. %What did you
                                %want to prove? - JJ
The left column has the
results when using five populations, and the right side shows the results for
ten. % Why only 5 and 10? - JJ
     % mmm I really do not have an answer for that, I was thinking about 
     % testing from 1 to 15, and also more workers - Mario

Charts show the execution time achieved by a certain number of workers.
Each box-plot indicates the time required for the execution of 15 function
instances, in seconds.  Each row has a different scale on the y-axis, to
compensate for the dimension increments. It is important to notice that the
execution time can also be affected by the algorithm finding a solution with an
error smaller than the threshold, before completing all the FEs.

As expected, in most cases, increasing the number of workers reduces the time
required for the execution of an instance. Mainly when using the same parameters
(number of populations and number of workers). But, when we compare the two
options, we find some subtle differences: When using one or two
workers, using 5 populations is better in most cases. A 
reason for this could be that the migration of individuals takes more time to
reach the worker. For instance, when a single worker has 10 populations, only
after processing every one of them, it will start receiving mixed populations.  Although, in
all cases, and even a single worker, populations are heterogeneous, this
advantage takes more time to propagate. With four workers, almost the same
results can be achieved regardless of our selection between five or ten
populations. Also, there is not much difference between 4 and 8 workers, when
using five populations. In general, we can observe that keeping a ratio close to
1:1 between the number of workers and populations  better exploits the computing
resources available.
% I don't really understand this too well... JJ
% Also, it might just be that populations are waiting on the queue to
% be processed if there are more sub-populations than workers. The
% fact that 10 subpopulations take more with 1w than 5 subpopulations
% might be due to late startup as you say above, but you need to
% explain it better.

% I was thinking also in terms of the heterogeneous populations.
% If population 1 to 6 are exploiting, and 7 to 9 are exploring
% they will take longer to catchup. 
% While with more workers this happens almost at the start
% If it makes sense, I will try to explain it better tomorrow with a 
% clear head - Mario

% Actually, what I see is no significant difference, or diminishing
% difference, between 5 and 10 sub-populations as the number of
% workers increases, and as long as the number of workers is less than
% the number of subpopulations.

% I think that was the point, but not too much less for instance 1w and 10pop
% Also I see a very big jump between 1w and 2w - Mario

>>>>

We present the results of the experiments in this section.
% The effect of a
% different number of populations on execution time is reported in  Section
% \ref{sec:exp1results}.
% The results of the performance comparison between homogeneous and
% heterogeneous search strategies are presented in Section \ref{sec:exp2results}.     
%
\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{sp_size_worker}
    \caption{Time required, in seconds, to complete an instance of the function, with different number of workers and populations sizes }
    \label{fig:spworker}
\end{figure}
%
%
% \subsection{Experiment 1}
% \label{sec:exp1results}
%
with results summarized in Figure \ref{fig:spworker}.  The
experiment consisted of executing the same benchmark on EvoSwarm twice but
changing the number of populations used in each run. This is mainly
intended as a showcase of the flexibility of the tool presented in
this paper, but we also wanted to test how the number of workers
affected the time needed to reach a solution. Adding more populations
might add more overhead, so this might have a noticeable effect.
The left column in that figure has the
results when using five populations, and the right side shows the results for
ten. In principle the main issue is how to find the right number of
workers for a particular problem, but in general we wanted to check
the communication overhead when the size of the problem increases and
how the presented framework reacted to it. % Added some explanation
% here - JJ

Charts show the execution time achieved by a certain number of
workers, for every problem dimension, which, besides, changes the
population size.
Each box-plot indicates the time required for the execution of 15 function
instances, in seconds.  Each row has a different scale on the y-axis, to
compensate for the dimension increments. It is important to notice that the
execution time can also be affected by the algorithm finding a solution with an
error smaller than the threshold, before completing all the FEs.

As expected, and in most cases, increasing the number of workers reduces the time
required for the execution of an instance, mainly when using the same parameters
(number of populations and number of workers), that is, performance
scales with the number of workers for a fixed problem and number of
sub-populations used. However, when we compare the two
options, we find some subtle differences: When using one or two
workers, using 5 populations takes less time in most cases. A 
reason for this could be that the migration of individuals takes more time to
reach the worker. For instance, when a single worker has 10 populations, only
after processing every one of them, it will start receiving mixed populations.  Although, in
all cases, and even a single worker, populations are heterogeneous, this
advantage takes more time to propagate. With four workers, almost the same
results can be achieved regardless of our selection between five or ten
populations. Also, there is not much difference between 4 and 8 workers, when
using five populations, which might indicate that some workers are
idling while populations are being processed by other workers. In
general, we can observe that keeping a ratio close to 1:1, but a bit
better than that, between the number of workers and populations, better exploits the computing
resources available. This also decreases the relation between the
overhead needed to migrate and process incoming and outgoing
population and the total computing time: more populations will need
more time to do that.

Another additional issue is related to the algorithm itself. Many more
workers than populations will have a wider range of states of
evolution in the experiment; populations that will be mixed will have
very different fitness states, which means that a part of them will
probably have been not so useful, following the intermediate
disturbance hypothesis. When the number of workers is increased, the
difference between the state of evolution of populations will be
intermediate, leading to better results; eventually, when there are as
many workers as populations, the state will be quite similar, which is
shown with a lower gap in the time won adding more workers.
